Modeling Liver with Glisson's Capsule in Real-Time Medical Simulations

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
ADD: some reformulations in section 3 and 4
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@581 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\maketitle
\begin{abstract}
  \TODO{}
The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, 
each having different mechanical properties.

@@

The computation of elastic stiffness matrix follows the common derivation
%
\begin{eqnarray}
  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
\end{eqnarray}
%
where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
matrix, $h$ is the thickness and $A$ area of the element. In the previous
\eqref{mem2} follows from the fact that we assume constant thickness of the
element and \eqref{mem3} follows from the fact that the strain-displacement
matrix is constant in our case. The strain-displacement matrix for the CST
element can be expressed as:
%
\begin{equation}
  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
  \end{bmatrix}
\end{equation}
%
The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
The reader can refer to the respective literature for more thorough
description~\cite{Felippa2003}.

@@

Using directly the boundary of the tetrahedral mesh does not only solve the
problem of building the surface mesh, but has one more advantage: the nodes
of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled
together.
% and solved as one system.
%
%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
%Without the loss of generality we can assume the tetrahedron consists of
%nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
%and $p_3$. We can reorder the degrees of freedom so that the stiffness
%matrix $\Mat{K}^t$ for the tetrahedron can be written as:
%%
%\begin{equation}
%  \Mat{K}^t = \left[\begin{array}{c|c}
%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%      \hline
%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
%  \end{array}\right]
%\end{equation}
%%
%Then the assembled stiffness matrix for the element is:
%%
%\begin{equation}
%  \Mat{K} = \left[\begin{array}{c|c}
%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%      \hline
%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
%  \end{array}\right]
%  +
%  \left[\begin{array}{c|c}
%      \Mat{K}^m & 0 \\
%      \hline
%      0 & 0 \\
%  \end{array}\right]
%\end{equation}
%%
%Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.

Alternatively, one can use a mechanical coupling similar to the one used in
\cite{Peterlik2012} to be able to use an arbitrary surface mesh. Nevertheless,
for conforming triangular mesh both methods lead to the same solution.

@@

\section{Methods and Simulation} %{{{

The model presented in the previous section was implemented in 
SOFA\footnote{www.sofa-framework.org} and a set of
numerical simulations was performed.
In this section we provide comparison of local deformations with the
results reported in literature to validate the method.
%Second, to show that in spite of its very small thickness the membrane cannot be
%neglected even in the context of global deformations and its overall
%stiffness plays an important role, the model of
%the complete liver was subjected to global deformations.

@@

A good example of such a local deformation is the aspiration test
where the response of liver exposed locally to a negative pressure is measured.
The aspiration device consists of a tube 1\,cm in diameter and allows
control of internal pressure in the tube. The test is performed by
attaching the tube to the tissue and measuring the tissue response. We
set up a simulation in SOFA to reproduce the experiment (see
Fig.~\ref{fig-aspiration1}: we meshed a 15$\times$15$\times$15\,mm$^3$ 
cube representing the tissue 
resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
and applied a pressure of 3\,kPa inside the tube. The interaction between the tube and 
tissue was modelled with friction contact.

In the simulations we used material properties for a pig liver reported in the literature: for the 
Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2011}.
Since the values for parenchyma elasticity quite vary (usually between 2\,kPa to 5\,kPa), we perform the simulation 
with $E_p$=3.5\,kPa.

\section{Results and Discussion} %{{{

@@

In~\cite{Hollenstein2006} it is reported that the Glisson's
capsule has a non-negligible influence on the response of the tissue and modeling
only the parenchyma leads to an overestimation of the deformation by a factor of 2 to 3. 
In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
test cube are presented showing significantly larger deformation in the model without capsule.
We can indeed observere that the deformation without the capsule is indeed overestimated approximately by a factor of 2. This is in perfect agreement with the results reported in the paper.% ~\cite{Hollenstein2006}

>>>>

\maketitle
\begin{abstract}
The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, 
each having different mechanical properties.

@@

The computation of elastic stiffness matrix follows the common derivation
%
\begin{eqnarray}
  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
\end{eqnarray}
%
where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
matrix, $h$ is the thickness and $A$ area of the element. In the previous
\eqref{mem2} follows from the fact that we assume constant thickness of the
element and \eqref{mem3} follows from the fact that the strain-displacement
matrix is constant in our case. The strain-displacement matrix for the CST
element can be expressed as:
%
\begin{equation}
  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
  \end{bmatrix}
\end{equation}
%
The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
description.

@@

Using directly the boundary of the tetrahedral mesh does not only solve the
problem of building the surface mesh, but has one more advantage: the nodes
of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled together.
% and solved as one system.
%
%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
Without the loss of generality we can assume the tetrahedron consists of
nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
and $p_3$. We can reorder the degrees of freedom so that the stiffness
matrix $\Mat{K}^t$ for the tetrahedron can be written as:
%
\begin{equation}
  \Mat{K}^t = \left[\begin{array}{c|c}
      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
      \hline
      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
  \end{array}\right]
\end{equation}
%
Then the assembled stiffness matrix for the element is:
%
\begin{equation}
  \Mat{K} = \left[\begin{array}{c|c}
      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
      \hline
      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
  \end{array}\right]
  +
  \left[\begin{array}{c|c}
      \Mat{K}^m & 0 \\
      \hline
      0 & 0 \\
  \end{array}\right]
\end{equation}
%
Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.

The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.

%Alternatively, one can use a mechanical coupling similar to the one used in
%\cite{Peterlik2012} to be able to use an arbitrary surface mesh. Nevertheless,
%for conforming triangular mesh both methods lead to the same solution.

@@

\section{Methods and Simulation} %{{{
The model presented in the previous section was implemented in 
SOFA\footnote{www.sofa-framework.org} and a set of
numerical simulations was performed.
In this section we provide comparison of local deformations with the
results reported in literature to validate the method.
%Second, to show that in spite of its very small thickness the membrane cannot be
%neglected even in the context of global deformations and its overall
%stiffness plays an important role, the model of
%the complete liver was subjected to global deformations.

@@

A good example of such a local deformation is the aspiration test
where the response of liver exposed locally to a negative pressure is measured.
The aspiration device consists of a tube having 1\,cm in diameter and allows to
control the pressure inside the tube. The test is performed by
attaching the tube to the tissue and measuring the tissue response. We
set up a simulation in SOFA to reproduce the experiment (see
Fig.~\ref{fig-aspiration1}: we meshed a 15$\times$15$\times$15\,mm$^3$ 
cube representing the tissue resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
and applied a pressure of 3\,kPa inside the tube.

Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
which allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
objects (see~\cite{} for details). Since NLCP requires explicitly the computation of compliance matrix which 
is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 
solve the system and compute the inverse matrices.

\section{Results and Discussion} %{{{
As the first step, we evaluate the influence of the Glissons's capsule on the mechanical response of the tissue during
the aspiration test. 
In~\cite{Hollenstein2006} it is reported that the influence of the capsule is significant, as modeling 
only the parenchyma without considering the membrane results in overestimation of the deformation by a factor of 2 to 3. 
In order to verify this observation, we performed a simulation using capsule parameters measured experimentally on 
ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
$t_c$=20\,$\mu$. 
Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.

@@

In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capculs is overestimated by a factor 
close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
